---
title: "Analysis of contextualized sensorimotor norms"
author: "Sean Trott"
date: "January 19, 2026"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "png")
```


```{r include=FALSE}
library(tidyverse)
library(lme4)
library(ggridges)
library(broom.mixed)
library(lmerTest)
library(ggcorrplot)
library(kableExtra)
```

# Introduction

In this document, we analyze the **contextualized sensorimotor norms**: judgments about the strength of different sensorimotor dimensions of ambiguous words, in context.

We use these norms in several analyses:

1. First, we compare them to the corresponding dimensions for the "decontextualized" Lancaster sensorimotor norms. 
2. Second, we ask whether the **dominance** of a word sense is correlated with its sensorimotor strength, i.e., whether more concrete meanings tend to be rated as more dominant.  
3. Third, we ask whether the **sensorimotor distance** between two contexts of use predicts judgments of how **related** those meanings are, above and beyond their distributional similarity and whether or not they belong to the same sense.
4. Fourth, we use **sensorimotor distance** to predict behavior on a primed sensibility judgment task.



# Characterizing dimensions

First, load the data.

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/cs_norms/src/analysis")
df_contextualized_meanings = read_csv("../../data/processed/contextualized_sensorimotor_norms.csv")
nrow(df_contextualized_meanings)
```


## Visualizing distributions

```{r different_dimensions}
df_contextualized_meanings_long = df_contextualized_meanings %>%
  pivot_longer(cols = c(Vision.M, Hearing.M, Olfaction.M,
                        Taste.M, Interoception.M, Touch.M,
                        Mouth_throat.M, Head.M, Torso.M,
                        Hand_arm.M, Foot_leg.M),
               names_to = "Dimension",
               values_to = "Strength") %>%
  mutate(Dimension = gsub('.M', '', Dimension))

df_contextualized_meanings_long %>%
  ggplot(aes(x = reorder(Dimension, Strength),
             y = Strength)) +
  geom_violin() +
  geom_jitter(alpha = .1,
              width = .1) +
  coord_flip() +
  labs(y = "Sensorimotor strength",
       x = "Dimension") +
  theme_bw() +
  theme(text = element_text(size=20))


```


## Correlations across dimensions

```{r corr_matrix}
columns = df_contextualized_meanings %>%
  mutate(Vision = Vision.M,
         Hearing = Hearing.M, 
         Olfaction = Olfaction.M,
         Taste = Taste.M, 
         Interoception = Interoception.M, 
         Touch = Touch.M,
         Mouth_throat = Mouth_throat.M,
         Head = Head.M,
         Torso = Torso.M,
         Hand_arm = Hand_arm.M, 
         Foot_leg = Foot_leg.M) %>%
  select(Vision, Hearing, Olfaction,
         Taste, Interoception, Touch,
         Mouth_throat, Head, Torso,
         Hand_arm, Foot_leg)
cors = cor(columns)

# cors[lower.tri(cors, diag=TRUE)] <- 0


# Plot the correlation matrix
ggcorrplot(cors, 
           hc.order = FALSE,
           # method = "square",
           type = "upper") +
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10)
  )

ggcorrplot(cors, hc.order = FALSE, type = "upper",
                      lab = TRUE, lab_size = 2.5) +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))




```


# Predicting dominance

## Load and merge data

Load the item-level means for the sensorimotor norms.

```{r}
df_contextualized_meanings = read_csv("../../data/processed/contextualized_sensorimotor_norms_with_ls.csv")
nrow(df_contextualized_meanings)
```


Load the dominance norms.

```{r}
df_dominance = read_csv("../../data/processed/dominance_norms_with_order.csv")

## Determine the specific sense/meaning of the righthand context
df_dominance = df_dominance %>%
  mutate(context = substr(version_with_order, 6, 9)) 

## Now group by that righthand context to get relative dominance of that meaning
df_dominance_individual = df_dominance %>%
  group_by(word, context) %>%
  summarise(dominance = mean(dominance_right))
nrow(df_dominance_individual)

```

Merge the dominance and sensorimotor norms data.

```{r}
df_dom_plus_sm = df_contextualized_meanings %>%
  inner_join(df_dominance_individual)
nrow(df_dom_plus_sm)
```


We also load and merge the Lancaster norms, as a control.

```{r}
df_lancaster <- read_csv("../../data/lexical/lancaster_norms.csv") %>%
  mutate(word = tolower(Word)) %>%
  rename(
    Foot_leg.SD.LS = Foot_leg.SD,
    Hand_arm.SD.LS = Hand_arm.SD,
    Head.SD.LS = Head.SD,
    Torso.SD.LS = Torso.SD
  )

df_dom_plus_sm <- df_dom_plus_sm %>%
  inner_join(df_lancaster)

nrow(df_dom_plus_sm)

```


## Calculating contextualized sensorimotor strength

Based on Lynott et al (2019), we **contextualized sensorimotor strength** as the *maximum* strength across all the dimensions.

```{r sm_strength_overall}
df_dom_plus_sm = df_dom_plus_sm %>%
  rowwise() %>%
  mutate(max_strength = max(
    c(
      ## Modalities
      Vision.M,
      Hearing.M,
      Olfaction.M,
      Touch.M,
      Taste.M, 
      Interoception.M,
      ## Effectors
      Head.M,
      Mouth_throat.M,
      Torso.M,
      Hand_arm.M,
      Foot_leg.M
    )
  ),
  minkowski3_strength = sum(c(Vision.M, Hearing.M, Olfaction.M, Touch.M, Taste.M,
                               Interoception.M, Head.M, Mouth_throat.M, Torso.M,
                               Hand_arm.M, Foot_leg.M)^3)^(1/3),
  max_perceptual_strength = max(
    c(
      ## Modalities
      Vision.M,
      Hearing.M,
      Olfaction.M,
      Touch.M,
      Taste.M, 
      Interoception.M
    )
  ),
  max_action_strength = max(
    c(
      ## Effectors
      Head.M,
      Mouth_throat.M,
      Torso.M,
      Hand_arm.M,
      Foot_leg.M
    )
  )
  ) %>%
  ungroup()

df_dom_plus_sm %>%
  ggplot(aes(x = Max_strength.sensorimotor,
             y = max_strength)) +
  geom_point(alpha = .5) +
  labs(y = "Maximum Contextualized Strength",
       x = "Maximum Strength (Lancaster)") +
  theme_bw()

cor.test(df_dom_plus_sm$Max_strength.sensorimotor,
         df_dom_plus_sm$max_strength)

cor.test(df_dom_plus_sm$max_strength,
         df_dom_plus_sm$minkowski3_strength)
```


## Does contextualized sensorimotor strength predict dominance?

The answer is **yes**: contexts with a higher *maximum* sensorimotor strength also tend to be rated as more *dominant*.

Notably, this is true above and beyond the *decontextualized* ratings of sensorimotor strength for a given word.

```{r strength_dominance}
df_dom_plus_sm %>%
  ggplot(aes(x = max_strength,
             y = dominance)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "lm") +
  labs(x = "Maximum Contextualized Strength",
       y = "Dominance") +
  theme_minimal() +
  theme(text = element_text(size=16))

  
df_dom_plus_sm %>%
  ggplot(aes(x = minkowski3_strength,
             y = dominance)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "lm") +
  labs(x = "Aggregate Sensorimotor Strength (Minkowski)",
       y = "Dominance") +
  theme_minimal() +
  theme(text = element_text(size=16))

model_full = lmer(data = df_dom_plus_sm,
                dominance ~ 
                  max_strength +
                  Max_strength.sensorimotor + Minkowski3.sensorimotor +
                  (1 | word),
                REML = FALSE)

model_reduced = lmer(data = df_dom_plus_sm,
                dominance ~ 
                  # max_strength + 
                  Max_strength.sensorimotor + Minkowski3.sensorimotor +
                  (1 | word),
                REML = FALSE)
summary(model_full)
anova(model_full, model_reduced)




model_full = lmer(data = df_dom_plus_sm,
                dominance ~ 
                  minkowski3_strength +
                  Max_strength.sensorimotor + Minkowski3.sensorimotor +
                  (1 | word),
                REML = FALSE)

model_reduced = lmer(data = df_dom_plus_sm,
                dominance ~ 
                  # max_strength + 
                  Max_strength.sensorimotor + Minkowski3.sensorimotor +
                  (1 | word),
                REML = FALSE)
summary(model_full)
anova(model_full, model_reduced)

```

### Examples

```{r}
### Top 10
df_dom_plus_sm %>%
  arrange(desc(minkowski3_strength)) %>%
  select(word, sentence, minkowski3_strength, dominance) %>%
  head(10)

### Bottom 10
df_dom_plus_sm %>%
  arrange(minkowski3_strength) %>%
  select(word, sentence, minkowski3_strength, dominance) %>%
  head(10)

### Top 10
df_dom_plus_sm %>%
  arrange(desc(max_strength)) %>%
  select(word, sentence, max_strength, dominance) %>%
  head(10)

### Bottom 10
df_dom_plus_sm %>%
  arrange(max_strength) %>%
  select(word, sentence, max_strength, dominance) %>%
  head(10)
```


### Which dimension best predicts dominance?

Here, we ask whether specific dimensions are particularly correlated with sense dominance.

```{r specific_dimensions_dominance}
features <- c("Vision.M", "Hearing.M", "Olfaction.M", "Touch.M", "Taste.M", 
              "Interoception.M", "Head.M", "Mouth_throat.M", "Torso.M", 
              "Hand_arm.M", "Foot_leg.M")

# Models with each feature + baseline covariate
r2_results <- map_dfr(features, function(feat) {
  
  formula <- as.formula(paste0("dominance ~ Max_strength.sensorimotor + ", feat))
  model <- lm(formula, data = df_dom_plus_sm)
  
  tibble(
    feature = feat,
    R2 = summary(model)$r.squared,
    beta = coef(model)[3],  # coefficient for the feature (3rd term now)
    p_value = summary(model)$coefficients[3, 4]
  )
})

# Baseline model (just Max_strength.sensorimotor)
baseline_model <- lm(dominance ~ Max_strength.sensorimotor, data = df_dom_plus_sm)

baseline_row <- tibble(
  feature = "Baseline",
  R2 = summary(baseline_model)$r.squared,
  beta = NA,
  p_value = NA
)

# Combine
r2_results_with_baseline <- bind_rows(baseline_row, r2_results)

# Plot
r2_results_with_baseline %>%
  mutate(feature = str_remove(feature, "\\.M$"),
         feature = str_replace(feature, "_", "/"),
         sig = ifelse(p_value < .05, "*", ""),
         is_baseline = feature == "Baseline",
         feature = fct_reorder(feature, R2)) %>%
  ggplot(aes(x = R2, y = feature, fill = is_baseline)) +
  geom_col() +
  geom_text(aes(label = sig), hjust = -0.5, size = 6, na.rm = TRUE) +
  scale_fill_manual(values = c("grey40", "steelblue"), guide = "none") +
  labs(x = "R²", y = NULL, title = "Predicting Sense Dominance") +
  theme_minimal() +
  theme(text = element_text(size = 16))


cor.test(df_dom_plus_sm$Touch.M,
         df_dom_plus_sm$dominance)

cor.test(df_dom_plus_sm$Vision.M,
         df_dom_plus_sm$dominance)
```


It looks like vision and touch are especially strong predictors. Which items drive this, e.g., for touch? 

```{r touch_dominance}
by_sense <- df_dom_plus_sm %>%
  mutate(sense = str_extract(context, "M[12]")) %>%
  group_by(word, sense) %>%
  summarize(
    mean_touch = mean(Touch.M),
    mean_vision = mean(Vision.M),
    mean_dom = mean(dominance),
    .groups = "drop"
  )

# Now 2 rows per word—compute within-word difference
sense_diffs <- by_sense %>%
  pivot_wider(names_from = sense, 
              values_from = c(mean_touch, mean_vision, mean_dom)) %>%
  mutate(
    touch_diff = mean_touch_M1 - mean_touch_M2,
    vision_diff = mean_vision_M1 - mean_vision_M2,
    dom_diff = mean_dom_M1 - mean_dom_M2
  )

sense_diffs %>%
  mutate(abs_touch_diff = abs(touch_diff)) %>%
  arrange(desc(abs_touch_diff)) %>%
  select(word, abs_touch_diff, dom_diff) %>%
  head(5)

sense_diffs %>%
  mutate(abs_vision_diff = abs(vision_diff)) %>%
  arrange(desc(abs_vision_diff)) %>%
  select(word, abs_vision_diff, dom_diff) %>%
  head(5)

# Does the sense with higher Touch or Vision tend to be more dominant?
cor.test(sense_diffs$touch_diff, sense_diffs$dom_diff)
cor.test(sense_diffs$vision_diff, sense_diffs$dom_diff)

```




# How much does context add?

Here, we inspect how much information the CS norms provide, both relative to the LS norms and just overall (i.e., how much context pushes sensorimotor vectors around).


## Comparing dimensions to Lancaster

Now, we compare each dimension to the LS Norms.

```{r ls_deviation_market}
df_diffs_raw <- df_dom_plus_sm %>%
  mutate(vision_diff       = (Vision.M       - Visual.mean),
         auditory_diff     = (Hearing.M      - Auditory.mean),
         intero_diff       = (Interoception.M - Interoceptive.mean),
         olfactory_diff    = (Olfaction.M    - Olfactory.mean),
         touch_diff        = (Touch.M        - Haptic.mean),
         taste_diff        = (Taste.M        - Gustatory.mean),
         torso_diff        = (Torso.M        - Torso.mean),
         hand_arm_diff     = (Hand_arm.M     - Hand_arm.mean),
         foot_leg_diff     = (Foot_leg.M     - Foot_leg.mean),
         head_diff         = (Head.M         - Head.mean),
         mouth_throat_diff = (Mouth_throat.M - Mouth.mean)) %>%
  pivot_longer(cols = ends_with("_diff"),
               names_to = "Dimension", values_to = "Diff") %>%
  mutate(Dimension = gsub('_diff', '', Dimension),
         Dimension = case_when(
           Dimension == "intero" ~ "Interoception",
           Dimension == "auditory" ~ "Hearing",
           Dimension == "olfactory" ~ "Olfaction",
           TRUE ~ str_to_title(Dimension)
         ))
df_diffs_raw$Dimension = factor(df_diffs_raw$Dimension,
                            levels = rev(c(
                              'Vision',
                              'Hearing',
                              'Olfaction',
                              'Taste',
                              'Interoception',
                              'Touch',
                              'Mouth_throat',
                              'Head',
                              'Torso',
                              'Hand_arm',
                              'Foot_leg'
                            )))






df_diffs_raw %>%
  filter(word == "market") %>%
  ggplot(aes(x = Dimension,
             y = Diff,
             fill = Dimension)) +
  geom_bar(stat = "summary") + 
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_bw() +
  coord_flip() +
  labs(x = "Dimension",
       y = "Deviation from Lancaster Norms") +
  scale_fill_manual(values = viridisLite::viridis(11, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~sentence) +
  theme(text = element_text(size=16)) +
  guides(fill = FALSE)



df_paired_market <- df_dom_plus_sm %>%
  filter(word == "market") %>%
  select(sentence,
         Vision.M, Hearing.M, Olfaction.M, Taste.M, Interoception.M, Touch.M,
         Mouth_throat.M, Head.M, Torso.M, Hand_arm.M, Foot_leg.M,
         Visual.mean, Auditory.mean, Olfactory.mean, Gustatory.mean,
         Interoceptive.mean, Haptic.mean, Mouth.mean, Head.mean,
         Torso.mean, Hand_arm.mean, Foot_leg.mean) %>%
  pivot_longer(cols = -sentence, names_to = "raw_dim", values_to = "Rating") %>%
  mutate(
    Source = case_when(
      str_detect(raw_dim, "\\.M$") ~ "CS Norms (contextualized)",
      TRUE ~ "LS Norms (decontextualized)"
    ),
    Dimension = case_when(
      raw_dim %in% c("Vision.M", "Visual.mean") ~ "Vision",
      raw_dim %in% c("Hearing.M", "Auditory.mean") ~ "Hearing",
      raw_dim %in% c("Olfaction.M", "Olfactory.mean") ~ "Olfaction",
      raw_dim %in% c("Taste.M", "Gustatory.mean") ~ "Taste",
      raw_dim %in% c("Interoception.M", "Interoceptive.mean") ~ "Interoception",
      raw_dim %in% c("Touch.M", "Haptic.mean") ~ "Touch",
      raw_dim %in% c("Mouth_throat.M", "Mouth.mean") ~ "Mouth_throat",
      raw_dim %in% c("Head.M", "Head.mean") ~ "Head",
      raw_dim %in% c("Torso.M", "Torso.mean") ~ "Torso",
      raw_dim %in% c("Hand_arm.M", "Hand_arm.mean") ~ "Hand_arm",
      raw_dim %in% c("Foot_leg.M", "Foot_leg.mean") ~ "Foot_leg"
    )
  )

df_paired_market$Dimension <- factor(df_paired_market$Dimension,
  levels = rev(c('Vision','Hearing','Olfaction','Taste','Interoception',
                 'Touch','Mouth_throat','Head','Torso','Hand_arm','Foot_leg')))

df_paired_market %>%
  ggplot(aes(x = Dimension, y = Rating, fill = Source)) +
  geom_bar(stat = "summary", position = position_dodge(width = 0.8), width = 0.7) +
  theme_bw() +
  coord_flip() +
  labs(x = "Dimension", y = "Sensorimotor Strength", fill = NULL) +
  scale_fill_manual(values = c("CS Norms (contextualized)" = "#2c7a7b",
                               "LS Norms (decontextualized)" = "#7c5e9c")) +
  facet_wrap(~sentence) +
  theme(text = element_text(size = 14), legend.position = "bottom")

```



### Distribution of deviations


```{r largest_dev}
df_diffs_raw %>%
  mutate(abs_diff = abs(Diff)) %>%
  summarise(mean_abs_diff = mean(abs_diff),
            median_abs_diff = median(abs_diff),
            sd_abs_diff = sd(abs_diff),
            min_abs_diff = min(abs_diff),
            max_abs_diff = max(abs_diff))

df_diffs_raw %>%
  mutate(abs_diff = abs(Diff)) %>%
  ggplot(aes(x = abs_diff)) +
  geom_histogram() +
  labs(x = "Absolute Deviation") +
  theme_minimal() +
  facet_wrap(~Dimension) +
   theme(text = element_text(size=16))


df_by_sentence_raw = df_diffs_raw %>%
  mutate(abs_diff = abs(Diff)) %>%
  group_by(word, sentence) %>%
  summarise(mean_abs_diff = mean(abs_diff),
            min_diff = min(abs_diff),
            max_diff = max(abs_diff)) %>%
  ungroup()

df_by_sentence_raw %>%
  summarise(mean_diff = mean(mean_abs_diff),
            median_abs_diff = median(mean_abs_diff),
            sd_abs_diff = sd(mean_abs_diff),
            min_abs_diff = min(mean_abs_diff),
            max_abs_diff = max(mean_abs_diff))


df_by_sentence_raw %>%
  ggplot(aes(x = mean_abs_diff)) +
  geom_histogram() +
  labs(x = "Mean Absolute Deviation (By Sentence)") +
  theme_minimal() +
   theme(text = element_text(size=16))

df_by_sentence_raw %>%
  ggplot(aes(x = max_diff)) +
  geom_histogram() +
  labs(x = "Max Absolute Deviation (By Sentence)") +
  theme_minimal() +
   theme(text = element_text(size=16))


df_by_sentence_raw %>%
  arrange(desc(max_diff)) %>%
  head(10)




df_by_word_raw = df_diffs_raw %>%
  mutate(abs_diff = abs(Diff)) %>%
  group_by(word) %>%
  summarise(mean_abs_diff = mean(abs_diff)) %>%
  ungroup()

df_by_word_raw %>%
  summarise(mean_diff = mean(mean_abs_diff),
            median_abs_diff = median(mean_abs_diff),
            sd_abs_diff = sd(mean_abs_diff),
            min_abs_diff = min(mean_abs_diff),
            max_abs_diff = max(mean_abs_diff))

  
df_diffs_raw %>%
  mutate(abs_diff = abs(Diff)) %>%
  group_by(word) %>%
  summarise(mean_abs_diff = mean(abs_diff),
            sd_abs_diff = sd(abs_diff),
            range_diff = max(abs_diff) - min(abs_diff)) %>%
  arrange(desc(mean_abs_diff)) %>%
  head(5)



```

### Individual words

```{r individual_words}

df_diffs_raw = df_diffs_raw %>%
  mutate(sense = str_extract(context, "M[12]"),         # "M1" or "M2"
         context_within_sense = str_extract(context, "[ab]"))  # "a" or "b"

df_diffs_raw %>%
  filter(word == "jam") %>%
  mutate(sentence = fct_reorder(sentence, as.numeric(factor(sense)))) %>%
  ggplot(aes(x = Dimension,
             y = Diff,
             fill = Dimension)) +
  geom_bar(stat = "summary") + 
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_bw() +
  coord_flip() +
  labs(x = "Dimension",
       y = "Deviation from Lancaster Norms") +
  scale_fill_manual(values = viridisLite::viridis(11, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~sentence) +
  theme(text = element_text(size=16)) +
  guides(fill = FALSE)

df_diffs_raw %>%
  mutate(sentence = fct_reorder(sentence, as.numeric(factor(sense)))) %>%
  filter(word == "punch") %>%
  ggplot(aes(x = Dimension,
             y = Diff,
             fill = Dimension)) +
  geom_bar(stat = "summary") + 
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_bw() +
  coord_flip() +
  labs(x = "Dimension",
       y = "Deviation from Lancaster Norms") +
  scale_fill_manual(values = viridisLite::viridis(11, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~sentence) +
  theme(text = element_text(size=16)) +
  guides(fill = FALSE)

df_diffs_raw %>%
  mutate(sentence = fct_reorder(sentence, as.numeric(factor(sense)))) %>%
  filter(word == "match") %>%
  ggplot(aes(x = Dimension,
             y = Diff,
             fill = Dimension)) +
  geom_bar(stat = "summary") + 
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_bw() +
  coord_flip() +
  labs(x = "Dimension",
       y = "Deviation from Lancaster Norms") +
  scale_fill_manual(values = viridisLite::viridis(11, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~sentence) +
  theme(text = element_text(size=16)) +
  guides(fill = FALSE)


df_diffs_raw %>%
  mutate(sentence = fct_reorder(sentence, as.numeric(factor(sense)))) %>%
  filter(word == "pen") %>%
  ggplot(aes(x = Dimension,
             y = Diff,
             fill = Dimension)) +
  geom_bar(stat = "summary") + 
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_bw() +
  coord_flip() +
  labs(x = "Dimension",
       y = "Deviation from Lancaster Norms") +
  scale_fill_manual(values = viridisLite::viridis(11, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~sentence) +
  theme(text = element_text(size=16)) +
  guides(fill = FALSE)


df_diffs_raw %>%
  mutate(sentence = fct_reorder(sentence, as.numeric(factor(sense)))) %>%
  filter(word == "band") %>%
  ggplot(aes(x = Dimension,
             y = Diff,
             fill = Dimension)) +
  geom_bar(stat = "summary") + 
  geom_vline(xintercept = 0, linetype = "dotted") +
  theme_bw() +
  coord_flip() +
  labs(x = "Dimension",
       y = "Deviation from Lancaster Norms") +
  scale_fill_manual(values = viridisLite::viridis(11, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~sentence) +
  theme(text = element_text(size=16)) +
  guides(fill = FALSE)

```


### Which dimensions have the largest deviations?

```{r}
df_diffs_raw %>%
  mutate(abs_diff = abs(Diff)) %>%
  group_by(Dimension) %>%
  summarise(mean_abs_diff = mean(abs_diff),
            sd_abs_diff = sd(abs_diff),
            range_diff = max(abs_diff) - min(abs_diff)) %>%
  arrange(desc(range_diff))
```

## Z-scored deviations

```{r zscored_diffs}
SD_FLOOR <- 0.1

df_diffs_z_floored <- df_dom_plus_sm %>%
  mutate(
    vision_diff       = (Vision.M       - Visual.mean)        / pmax(Visual.SD, SD_FLOOR),
    auditory_diff     = (Hearing.M      - Auditory.mean)      / pmax(Auditory.SD, SD_FLOOR),
    intero_diff       = (Interoception.M - Interoceptive.mean) / pmax(Interoceptive.SD, SD_FLOOR),
    olfactory_diff    = (Olfaction.M    - Olfactory.mean)     / pmax(Olfactory.SD, SD_FLOOR),
    touch_diff        = (Touch.M        - Haptic.mean)        / pmax(Haptic.SD, SD_FLOOR),
    taste_diff        = (Taste.M        - Gustatory.mean)     / pmax(Gustatory.SD, SD_FLOOR),
    torso_diff        = (Torso.M        - Torso.mean)         / pmax(Torso.SD.LS, SD_FLOOR),
    hand_arm_diff     = (Hand_arm.M     - Hand_arm.mean)      / pmax(Hand_arm.SD.LS, SD_FLOOR),
    foot_leg_diff     = (Foot_leg.M     - Foot_leg.mean)      / pmax(Foot_leg.SD.LS, SD_FLOOR),
    head_diff         = (Head.M         - Head.mean)          / pmax(Head.SD.LS, SD_FLOOR),
    mouth_throat_diff = (Mouth_throat.M - Mouth.mean)         / pmax(Mouth.SD, SD_FLOOR)
  ) %>%
  pivot_longer(cols = ends_with("_diff"),
               names_to = "Dimension", values_to = "Diff") %>%
  mutate(Dimension = gsub('_diff', '', Dimension),
         Dimension = case_when(
           Dimension == "intero" ~ "Interoception",
           Dimension == "auditory" ~ "Hearing",
           Dimension == "olfactory" ~ "Olfaction",
           TRUE ~ str_to_title(Dimension)
         )) %>%
  select(word, sentence, Dimension, Diff, context, dominance)




# Distribution
sentence_dev_floored <- df_diffs_z_floored %>%
  mutate(abs_diff = abs(Diff)) %>%
  group_by(word, sentence) %>%
  summarise(mean_abs_diff = mean(abs_diff, na.rm = TRUE),
            max_abs_diff = max(abs_diff),
            .groups = "drop")

sentence_dev_floored %>%
  ggplot(aes(x = mean_abs_diff)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  labs(x = "Mean Absolute Standardized Deviation (By Sentence)",
       y = "Count of sentences") +
  theme_minimal() +
  theme(text = element_text(size=16))

sentence_dev_floored %>%
  ggplot(aes(x = max_abs_diff)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  labs(x = "Max Absolute Standardized Deviation (By Sentence)",
       y = "Count of sentences") +
  theme_minimal() +
  theme(text = element_text(size=16))


### Overall % sentences more than one on average
sentence_dev_floored %>%
  mutate(greater_than_one = mean_abs_diff > 1) %>%
  summarise(prop_more_than_one = mean(greater_than_one),
            max_dev = max(mean_abs_diff))

### Overall % sentences more than one maximally
quantile(sentence_dev_floored$max_abs_diff, 
         probs = c(0.25, 0.5, 0.75, 0.9, 0.95, 0.99),
         na.rm = TRUE)


### Overall deviations by dimension
df_diffs_z_floored %>%
  mutate(abs_diff = abs(Diff)) %>%
  ggplot(aes(x = abs_diff)) +
  geom_histogram() +
  labs(x = "Standardized Absolute Deviation") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  theme_minimal() +
  facet_wrap(~Dimension) +
   theme(text = element_text(size=16))

### Overall % of contexts exceeding 1 SD
df_diffs_z_floored %>%
  mutate(abs_diff = abs(Diff)) %>%
  mutate(greater_than_one = abs_diff > 1) %>%
  summarise(prop_more_than_one = mean(greater_than_one),
            max_dev = max(abs_diff))

#### Additional summary statistics
df_diffs_z_floored %>%
  mutate(abs_diff = abs(Diff)) %>%
  summarise(mean_abs_diff = mean(abs_diff),
            median_abs_diff = median(abs_diff),
            sd_abs_diff = sd(abs_diff),
            min_abs_diff = min(abs_diff),
            max_abs_diff = max(abs_diff))

df_by_sentence_z_floored = df_diffs_z_floored %>%
  mutate(abs_diff = abs(Diff)) %>%
  group_by(word, sentence) %>%
  summarise(mean_abs_diff = mean(abs_diff)) %>%
  ungroup()

df_by_sentence_z_floored %>%
  summarise(mean_diff = mean(mean_abs_diff),
            median_abs_diff = median(mean_abs_diff),
            sd_abs_diff = sd(mean_abs_diff),
            min_abs_diff = min(mean_abs_diff),
            max_abs_diff = max(mean_abs_diff))

df_by_word_z_floored = df_diffs_z_floored %>%
  mutate(abs_diff = abs(Diff)) %>%
  group_by(word) %>%
  summarise(mean_abs_diff = mean(abs_diff)) %>%
  ungroup()

df_by_word_z_floored %>%
  summarise(mean_diff = mean(mean_abs_diff),
            median_abs_diff = median(mean_abs_diff),
            sd_abs_diff = sd(mean_abs_diff),
            min_abs_diff = min(mean_abs_diff),
            max_abs_diff = max(mean_abs_diff))

```


### Compare to raw differences

```{r}

# Raw (unstandardized) mean absolute deviation per sentence
sentence_dev_raw <- df_diffs_raw %>%
  mutate(abs_diff = abs(Diff)) %>%
  group_by(word, sentence, context) %>%
  summarise(mean_abs_diff = mean(abs_diff, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(version = "Raw")

# Standardized (floored) mean absolute deviation per sentence
sentence_dev_z <- df_diffs_z_floored %>%
  mutate(abs_diff = abs(Diff)) %>%
  group_by(word, sentence, context) %>%
  summarise(mean_abs_diff = mean(abs_diff, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(version = "Standardized")


# By sentence
sentence_compare <- sentence_dev_raw %>%
  select(word, sentence, context, raw = mean_abs_diff) %>%
  inner_join(sentence_dev_z %>%
               select(word, sentence, context, standardized = mean_abs_diff),
             by = c("word", "sentence", "context"))

cor.test(sentence_compare$raw, sentence_compare$standardized, method = "spearman")
cor.test(sentence_compare$raw, sentence_compare$standardized, method = "pearson")

sentence_compare %>%
  ggplot(aes(x = raw, y = standardized)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Raw mean absolute deviation",
       y = "Standardized mean absolute deviation",
       title = "By-sentence deviations") +
  theme_minimal() +
  theme(text = element_text(size=16))


```


## Within-word range

Here, we calculate the *range* of sensorimotor associatoins for each dimension across the contexts of use for each word.

### Across all contexts

```{r within_word_range}

df_rawc_with_norms = read_csv("../../data/processed/sentence_pairs_with_sensorimotor_distance.csv") %>%
  drop_na(sensorimotor_distance) %>%
  select(sensorimotor_distance, action_distance, perceptual_distance,
         word, same, ambiguity_type, sentence1, sentence2, mean_relatedness, Class, version)

nrow(df_rawc_with_norms)


word_dim_range <- df_dom_plus_sm %>%
  pivot_longer(cols = c(Vision.M, Hearing.M, Olfaction.M, Taste.M,
                        Interoception.M, Touch.M, Mouth_throat.M,
                        Head.M, Hand_arm.M, Foot_leg.M, Torso.M),
               names_to = "Dimension", values_to = "CS_value") %>%
  mutate(Dimension = gsub("\\.M$", "", Dimension)) %>%
  group_by(word, Dimension) %>%
  summarise(within_range = max(CS_value, na.rm = TRUE) - min(CS_value, na.rm = TRUE),
            .groups = "drop")

# Aggregate to word level (mean range across dimensions)
word_range <- word_dim_range %>%
  group_by(word) %>%
  summarise(mean_range = mean(within_range, na.rm = TRUE),
            max_range = max(within_range, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(desc(mean_range))

print(head(word_range, 15))
print(tail(word_range, 15))

# By dimension
word_dim_range %>%
  ggplot(aes(x = within_range)) +
  geom_histogram(bins = 30) +
  labs(x = "Within-Word Range of CS Ratings (across contexts)",
       y = "Count of words") +
  theme_minimal() +
  facet_wrap(~Dimension)

# Distribution
word_range %>%
  ggplot(aes(x = mean_range)) +
  geom_histogram(bins = 30) +
  labs(x = "Mean Within-Word Range of CS Ratings (across contexts)",
       y = "Count of words") +
  theme_minimal()

# Summary stats
word_range %>%
  summarise(mean = mean(mean_range),
            sd = sd(mean_range),
            median = median(mean_range),
            min = min(mean_range),
            max = max(mean_range))


# Get per-word ambiguity type
word_ambiguity <- df_rawc_with_norms %>%
  distinct(word, ambiguity_type)

# Merge with within-word range
word_range_with_amb <- word_range %>%
  inner_join(word_ambiguity, by = "word")

# Summary stats by ambiguity type
word_range_with_amb %>%
  group_by(ambiguity_type) %>%
  summarise(mean = mean(mean_range),
            sd = sd(mean_range),
            median = median(mean_range),
            n = n(),
            .groups = "drop")

# Test the difference
summary(lm(data = word_range_with_amb,
           mean_range ~ ambiguity_type))
t.test(mean_range ~ ambiguity_type, data = word_range_with_amb)


# Visualize
word_range_with_amb %>%
  ggplot(aes(x = ambiguity_type, y = mean_range, fill = ambiguity_type)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.15, alpha = 0.3) +
  labs(x = "Ambiguity Type",
       y = "Mean Within-Word Range of CS Ratings") +
  theme_minimal() +
  theme(legend.position = "none")


```

### Between vs. Within sense

```{r within_word_range_senses}
# Need sense info from context column (M1/M2)
word_range_decomposed <- df_dom_plus_sm %>%
  mutate(sense = str_extract(context, "M[12]")) %>%
  pivot_longer(cols = c(Vision.M, Hearing.M, Olfaction.M, Taste.M,
                        Interoception.M, Touch.M, Mouth_throat.M,
                        Head.M, Hand_arm.M, Foot_leg.M, Torso.M),
               names_to = "Dimension", values_to = "CS_value") %>%
  mutate(Dimension = gsub("\\.M$", "", Dimension)) %>%
  group_by(word, Dimension, sense) %>%
  summarise(sense_mean = mean(CS_value),
            sense_within_range = max(CS_value) - min(CS_value),
            .groups = "drop") %>%
  group_by(word, Dimension) %>%
  summarise(between_sense_range = max(sense_mean) - min(sense_mean),
            within_sense_range = mean(sense_within_range),
            .groups = "drop") %>%
  group_by(word) %>%
  summarise(mean_between_sense = mean(between_sense_range),
            mean_within_sense = mean(within_sense_range),
            max_between_sense = max(between_sense_range),
            max_within_sense = max(within_sense_range),
            .groups = "drop")


word_range_decomposed %>%
  inner_join(word_ambiguity, by = "word") %>%
  group_by(ambiguity_type) %>%
  summarise(between = mean(mean_between_sense),
            within = mean(mean_within_sense),,
            between_max = mean(max_between_sense),
            within_max = mean(max_within_sense),
            n = n())


# Test each separately
t.test(mean_between_sense ~ ambiguity_type, 
       data = word_range_decomposed %>% inner_join(word_ambiguity, by = "word"))
t.test(mean_within_sense ~ ambiguity_type,
       data = word_range_decomposed %>% inner_join(word_ambiguity, by = "word"))

# Test max
t.test(max_between_sense ~ ambiguity_type, 
       data = word_range_decomposed %>% inner_join(word_ambiguity, by = "word"))
t.test(max_within_sense ~ ambiguity_type,
       data = word_range_decomposed %>% inner_join(word_ambiguity, by = "word"))


word_range_decomposed_long = word_range_decomposed %>%
  inner_join(word_ambiguity, by = "word") %>%
  pivot_longer(cols = c(mean_between_sense, mean_within_sense),
               names_to = "comparison",
               values_to = "word_range")

summary(lmer(data = word_range_decomposed_long,
           word_range ~ comparison + (1|word)))

summary(lmer(data = word_range_decomposed_long,
           word_range ~ ambiguity_type*comparison + (1|word)))



word_range_decomposed_long %>%
  mutate(comparison = case_when(
    comparison == "mean_between_sense" ~ "Different Sense",
    comparison == "mean_within_sense" ~ "Same Sense"
  )) %>%
  ggplot(aes(x = word_range,
             y = ambiguity_type,
             fill = comparison)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color = NA,
                       alpha = 0.5, 
                       scale=0.85, 
                       stat="density") +
  labs(x = "Mean Within-Word Range of CS Ratings",
       y = NULL,
       fill = NULL) +
  theme_minimal() +
  scale_fill_manual(
    values = c("Same Sense" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "Different Sense" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) + 
  theme(text = element_text(size = 20))




word_range_decomposed_long = word_range_decomposed %>%
  inner_join(word_ambiguity, by = "word") %>%
  pivot_longer(cols = c(max_between_sense, max_within_sense),
               names_to = "comparison",
               values_to = "word_range")



summary(lmer(data = word_range_decomposed_long,
           word_range ~ comparison + (1|word)))

summary(lmer(data = word_range_decomposed_long,
           word_range ~ ambiguity_type*comparison + (1|word)))



word_range_decomposed_long %>%
  mutate(comparison = case_when(
    comparison == "max_between_sense" ~ "Different Sense",
    comparison == "max_within_sense" ~ "Same Sense"
  )) %>%
  ggplot(aes(x = word_range,
             y = ambiguity_type,
             fill = comparison)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color = NA,
                       alpha = 0.5, 
                       scale=0.85, 
                       stat="density") +
  labs(x = "Max Within-Word Range of CS Ratings",
       y = NULL,
       fill = NULL) +
  theme_minimal() +
  scale_fill_manual(
    values = c("Same Sense" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "Different Sense" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) + 
  theme(text = element_text(size = 20))
```



# How does dominance relate to deviation from the LS Norms?

This question can in turn be decomposed into two questions:

First, are more dominant senses **closer** to the LS Norms overall? We might expect this to be the case if the LS Norms reflect the dominant sense; that is, when people rate the sensorimotor properties of a decontextualized word, they might be more likely to index properties associated with the most dominant contexts or meanings of that word.

And the answer is **yes**: more dominant senses are indeed more *similar* (less distant) from the Lancaster norm in terms of their sensorimotor profile.

```{r dominance_ls_deviation}
model_with_dominance = lmer(data = df_dom_plus_sm,
                  distance_to_lancaster ~ dominance + (1 | word),
                  REML = FALSE)

model_no_dominance = lmer(data = df_dom_plus_sm,
                  distance_to_lancaster ~ (1 | word),
                  REML = FALSE)

summary(model_with_dominance)
anova(model_with_dominance, model_no_dominance)

df_dom_plus_sm %>%
  ggplot(aes(x = dominance,
             y = distance_to_lancaster)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm") +
  labs(x = "Dominance",
       y = "Cosine Distance to Decontextualized LS Norm") +
  theme_bw()

cor.test(df_dom_plus_sm$dominance, df_dom_plus_sm$distance_to_lancaster)
```

And second: does dominance predict the **direction** of difference?

The earlier analysis of dominance suggests that more dominant senses are more concrete than less dominant senses. Thus, we might expect that more dominant senses are also more concrete on average than the decontextualized norms. 

We find that this is true: that is, more dominant senses are *more concrete* on average than the LS norm.

```{r}
df_diffs_avg = df_diffs_raw %>%
  group_by(word, sentence, context) %>%
  summarise(mean_diff = mean(Diff))


df_diffs_avg = df_diffs_avg %>%
  left_join(df_dom_plus_sm)

model_with_dominance = lmer(data = df_diffs_avg,
                  mean_diff ~ dominance + (1 | word),
                  REML = FALSE)

model_no_dominance = lmer(data = df_diffs_avg,
                  mean_diff ~ (1 | word),
                  REML = FALSE)

summary(model_with_dominance)
anova(model_with_dominance, model_no_dominance)

cor.test(df_diffs_avg$dominance, df_diffs_avg$mean_diff)

```



# Predicting relatedness 

Next, we ask about the **sensorimotor distance** between two sentence pairs, and whether it correlates both with `same/different sense` and the `mean_relatedness` judgments for those sentence pairs.

Here, we load a version of the dataset that also contains a *baseline* measure: the sensorimotor distance as calculated using a bag-of-words approach (i.e., using the original Lancaster Norms).

## Load data

```{r}
df_rawc_with_norms = read_csv("../../data/processed/sentence_pairs_with_sensorimotor_distance.csv") %>%
  drop_na(sensorimotor_distance) %>%
  select(sensorimotor_distance, action_distance, perceptual_distance,
         word, same, ambiguity_type, sentence1, sentence2, mean_relatedness, Class, version)

nrow(df_rawc_with_norms)

```



## Load English RAW-C data

```{r}
df_bert = read_csv("../../data/processed/models_english/rawc-distances_model-bert-base-uncased.csv") %>%
  mutate(Model = "BERT-base-uncased",
         Multilingual = "Monolingual")

df_bert_cased = read_csv("../../data/processed/models_english/rawc-distances_model-bert-base-cased.csv") %>%
  mutate(Model = "BERT-base-cased",
         Multilingual = "Monolingual")


df_xlm = read_csv("../../data/processed/models_english/rawc-distances_model-xlm-roberta-base.csv") %>%
  mutate(Model = "XLM-RoBERTa",
         Multilingual = "Multilingual")


df_ab1 = read_csv("../../data/processed/models_english/rawc-distances_model-albert-base-v1.csv") %>%
  mutate(Model = "ALBERT-base-v1",
         Multilingual = "Monolingual")

df_ab2 = read_csv("../../data/processed/models_english/rawc-distances_model-albert-base-v2.csv") %>%
  mutate(Model = "ALBERT-base-v2",
         Multilingual = "Monolingual")

df_al = read_csv("../../data/processed/models_english/rawc-distances_model-albert-large-v2.csv") %>%
  mutate(Model = "ALBERT-large-v2",
         Multilingual = "Monolingual")

df_axl = read_csv("../../data/processed/models_english/rawc-distances_model-albert-xlarge-v2.csv") %>%
  mutate(Model = "ALBERT-xlarge-v2",
         Multilingual = "Monolingual")

df_axxl = read_csv("../../data/processed/models_english/rawc-distances_model-albert-xxlarge-v2.csv")  %>%
  mutate(Model = "ALBERT-xxlarge-v2",
         Multilingual = "Monolingual")

df_rb = read_csv("../../data/processed/models_english/rawc-distances_model-roberta-base.csv") %>%
  mutate(Model = "RoBERTa-base",
         Multilingual = "Monolingual")

df_rl = read_csv("../../data/processed/models_english/rawc-distances_model-roberta-large.csv") %>%
  mutate(Model = "RoBERTa-large",
         Multilingual = "Monolingual")

df_db = read_csv("../../data/processed/models_english/rawc-distances_model-distilbert-base-uncased.csv") %>%
  mutate(Model = "DistilBERT",
         Multilingual = "Monolingual")


df_mb = read_csv("../../data/processed/models_english/rawc-distances_model-bert-base-multilingual-cased.csv") %>%
  mutate(Model = "Multilingual BERT",
         Multilingual = "Multilingual")



df_all = df_bert %>%
  bind_rows(df_bert_cased) %>%
  bind_rows(df_xlm) %>%
  bind_rows(df_ab1) %>%
  bind_rows(df_ab2) %>%
  bind_rows(df_al) %>%
  bind_rows(df_axl) %>%
  bind_rows(df_axxl) %>%
  bind_rows(df_rb) %>%
  bind_rows(df_rl) %>%
  bind_rows(df_db) %>%
  bind_rows(df_mb)


df_merged = df_rawc_with_norms %>%
  inner_join(df_all)
```


## Predicting same/different sense

### Sensorimotor distance and sense boundaries


```{r sm_sense}

df_rawc_with_norms = df_rawc_with_norms %>%
  mutate(Same = case_when(
    same == TRUE ~ "Same Sense",
    same == FALSE ~ "Different Sense"
  ))



df_rawc_with_norms %>%
  ggplot(aes(x = sensorimotor_distance,
             y = ambiguity_type,
             fill = Same)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color = NA,
                       alpha = 0.5, 
                       scale=0.85, 
                       stat="density") +
  labs(x = "Sensorimotor Distance",
       y = NULL,
       fill = NULL) +
  scale_fill_manual(
    values = c("Same Sense" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "Different Sense" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) + 
  theme_minimal() +
  theme(text = element_text(size = 20))



model_full = lmer(data = df_rawc_with_norms,
                  sensorimotor_distance ~ same +
                    (1 + same | word),
                  control=lmerControl(optimizer="bobyqa"),
                  REML = FALSE)

model_reduced = lmer(data = df_rawc_with_norms,
                  sensorimotor_distance ~ # same +
                    (1 + same | word),
                  control=lmerControl(optimizer="bobyqa"),
                  REML = FALSE)

summary(model_full)
anova(model_full, model_reduced)


model_full_at = lmer(data = df_rawc_with_norms,
                  sensorimotor_distance ~ same * ambiguity_type +
                    (1 + same | word),
                  control=lmerControl(optimizer="bobyqa"),
                  REML = FALSE)

model_reduced_at = lmer(data = df_rawc_with_norms,
                  sensorimotor_distance ~ same + ambiguity_type +
                    (1 + same | word),
                  control=lmerControl(optimizer="bobyqa"),
                  REML = FALSE)

summary(model_full_at)
anova(model_full_at, model_reduced_at)


df_rawc_with_norms %>%
  group_by(same, ambiguity_type) %>%
  summarise(mean = mean(sensorimotor_distance),
            sd = sd(sensorimotor_distance))


```


### Predicting sense boundaries

Here, we run the actual analysis:

```{r aic_sense_boundaries}


model_sm_only <- glm(same ~ sensorimotor_distance, 
                     data = df_rawc_with_norms, 
                     family = binomial)


aic_sm_baseline <- AIC(model_sm_only)

print(paste("AIC:", round(aic_sm_baseline, 1)))


### Get aic
aic_results <- df_merged %>%
  group_by(n_params, Model, Layer) %>%
  summarize(
    # BERT only
    aic_bert = {
      model <- glm(Same_sense ~ Distance, family = binomial)
      AIC(model)
    },
    # Combined
    aic_combined = {
      model <- glm(Same_sense ~ Distance + sensorimotor_distance, family = binomial)
      AIC(model)
    },
    aic_sm = {
      model <- glm(Same_sense ~ sensorimotor_distance, family = binomial)
      AIC(model)
    },
    .groups = "drop"
  ) %>%
  mutate(
    # AIC Delta
    aic_bert_vs_sm = aic_bert - aic_sm_baseline,
    aic_combined_vs_sm = aic_sm_baseline - aic_combined,
    aic_combined_vs_dist = aic_bert - aic_combined
  )

# ============================================================
# 3. Best layer per model
# ============================================================

best_layers <- aic_results %>%
  group_by(Model, n_params) %>%
  slice_min(aic_bert, n = 1) %>%
  select(Model, Layer, 
         aic_bert_vs_sm, aic_combined_vs_sm, aic_combined_vs_dist,
          aic_sm, aic_bert, aic_combined)


best_layers = best_layers %>%
  mutate(aic_bert_vs_sm2 = aic_bert - aic_sm_baseline)

print(best_layers)

## AIC difference from SM baseline

best_layers %>%
  select(Model, aic_bert_vs_sm, aic_combined_vs_sm) %>%
  pivot_longer(cols = c(aic_bert_vs_sm, aic_combined_vs_sm),
               names_to = "type", values_to = "aic_diff") %>%
  mutate(type = ifelse(type == "aic_bert_vs_sm", 
                       "Distributional only", 
                       "Distributional + Sensorimotor"),
         type = factor(type, levels = c("Distributional only", 
                                        "Distributional + Sensorimotor")),
         Model = fct_reorder(Model, aic_diff)) %>%
  ggplot(aes(x = aic_diff, y = Model, fill = type)) +
  geom_col(position = "dodge") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "ΔAIC vs. Sensorimotor baseline", 
       y = NULL,
       fill = NULL) +
  scale_fill_manual(values = c("Distributional only" = "gray60", 
                                "Distributional + Sensorimotor" = "steelblue")) +
  theme_minimal() +
  theme(text = element_text(size = 16),
        legend.position = "bottom")

best_layers %>%
  ungroup() %>%
  select(Model, aic_bert, aic_combined) %>%
  mutate(across(starts_with("aic"), round)) %>%
  kbl(col.names = c("Model", "Distributional", "Hybrid"),
      format = "latex",
      booktabs = TRUE,
      caption = "AIC values for predicting sense boundaries using Distributional Distance alone or both Distributional Distance and Sensorimotor Distance (AIC for Sensorimotor Distance was 707)",
      label = "aic") %>%
  kable_styling(latex_options = "hold_position")


aic_results %>%
  mutate(dist_better_than_sm = aic_bert_vs_sm > 4,
         hybrid_better_than_sm = aic_combined_vs_sm > 4,
         hybrid_better_than_dist = aic_combined_vs_dist > 4) %>%
  summarise(mean(dist_better_than_sm),
            mean(hybrid_better_than_sm),
            mean(hybrid_better_than_dist))

best_layers %>%
  ungroup() %>%
  mutate(dist_better_than_sm = aic_bert_vs_sm < -4,
         hybrid_better_than_sm = aic_combined_vs_sm > 4,
         hybrid_better_than_dist = aic_combined_vs_dist > 4) %>%
  summarise(mean(dist_better_than_sm),
            mean(hybrid_better_than_sm),
            mean(hybrid_better_than_dist))




### Visualize raw AIC
aic_summary_full <- best_layers %>%
  select(Model, Layer, aic_sm, aic_bert, aic_combined) %>%
  pivot_longer(cols = c(aic_sm, aic_bert, aic_combined),
               names_to = "type", values_to = "aic") %>%
  mutate(type = case_when(
    type == "aic_sm" ~ "Sensorimotor",
    type == "aic_bert" ~ "Distributional",
    type == "aic_combined" ~ "Hybrid"
  ),
  type = factor(type, levels = c("Sensorimotor", "Distributional", "Hybrid")))

# Get sensorimotor baseline (should be the same for all models)
aic_sm_baseline <- aic_summary_full %>%
  filter(type == "Sensorimotor") %>%
  pull(aic) %>%
  unique()

# Filter to just distributional and hybrid
aic_summary_no_sm <- aic_summary_full %>%
  filter(type != "Sensorimotor")

aic_summary_se_no_sm <- aic_summary_no_sm %>%
  group_by(type) %>%
  summarize(
    mean_aic = mean(aic),
    se_aic = sd(aic) / sqrt(n())
  )

ggplot() +
  # Sensorimotor baseline as dashed line
  geom_hline(yintercept = aic_sm_baseline, 
             linetype = "dashed", color = "coral", linewidth = 1) +
  # Raw points for each model/layer
  geom_jitter(data = aic_summary_no_sm, 
              aes(x = type, y = aic, color = type),
              alpha = 0.3, width = 0.1, size = 2) +
  # Mean + SE
  geom_point(data = aic_summary_se_no_sm, 
             aes(x = type, y = mean_aic, color = type),
             size = 5) +
  geom_errorbar(data = aic_summary_se_no_sm, 
                aes(x = type, ymin = mean_aic - se_aic, 
                    ymax = mean_aic + se_aic, color = type), 
                width = 0.15, linewidth = 1.2) +
  annotate("text", x = 1.5, y = aic_sm_baseline + 30, 
           label = "Sensorimotor", color = "coral", size = 5)+
  labs(x = NULL, 
       y = "AIC (lower is better)",
       title = "Predicting Sense Boundary") +
  scale_color_manual(values = c("Distributional" = "gray60", 
                                "Hybrid" = "steelblue")) +
  theme_minimal() +
  theme(text = element_text(size = 16),
        legend.position = "none")

```


## Predicting relatedness

### Comparing to each distributional relatedness measure

```{r relatedness_comparison}

df_rawc_with_norms %>%
  ggplot(aes(x = sensorimotor_distance,
            y = mean_relatedness,
            color = Same)) +
  geom_point(alpha = .5) +
  scale_color_manual(
    values = c("Same Sense" = viridisLite::viridis(2, option = "mako", 
                                                   begin = 0.8, end = 0.15)[1],
               "Different Sense" = viridisLite::viridis(2, option = "mako", 
                                                        begin = 0.8, end = 0.15)[2])
  ) +
  theme_minimal() +
  labs(x = "Sensorimotor Distance",
       y = "Mean Relatedness",
       color = "") +
  theme(text = element_text(size = 15),
      legend.position="bottom")
  

cor(df_rawc_with_norms$sensorimotor_distance,
            df_rawc_with_norms$mean_relatedness)

```


### Comparing AIC

```{r aic_relatedness}

# ============================================================
# 2. For each model/layer: compute R² and AIC, difference from SM baseline
# ============================================================

relatedness_results <- df_merged %>%
  group_by(Model, Layer, n_params, Multilingual) %>%
  summarize(
    # Sensorimotor only
    r2_sm = {
      model <- lm(mean_relatedness ~ sensorimotor_distance)
      summary(model)$r.squared
    },
    aic_sm = {
      model <- lm(mean_relatedness ~ sensorimotor_distance)
      AIC(model)
    },
    # BERT only
    r2_bert = {
      model <- lm(mean_relatedness ~ Distance)
      summary(model)$r.squared
    },
    aic_bert = {
      model <- lm(mean_relatedness ~ Distance)
      AIC(model)
    },
    # Combined
    r2_combined = {
      model <- lm(mean_relatedness ~ Distance + sensorimotor_distance)
      summary(model)$r.squared
    },
    aic_combined = {
      model <- lm(mean_relatedness ~ Distance + sensorimotor_distance)
      AIC(model)
    },
    .groups = "drop"
  ) %>%
  mutate(
    aic_bert_vs_sm = aic_bert - aic_sm,
    aic_combined_vs_sm = aic_sm - aic_combined,
    aic_combined_vs_dist = aic_bert - aic_combined
  )

# ============================================================
# 3. Best layer per model
# ============================================================

best_layers <- relatedness_results %>%
  group_by(Model, n_params, Multilingual) %>%
  slice_max(r2_bert, n = 1) %>%
  select(Model, Layer, n_params, Multilingual,
         # r2_sm, r2_bert, r2_combined,
         aic_bert_vs_sm, aic_sm, aic_bert, aic_combined,
         aic_combined_vs_sm, aic_combined_vs_dist)

print(best_layers)

best_layers %>%
  ungroup() %>%
  select(Model, aic_bert, aic_combined) %>%
  mutate(across(starts_with("aic"), round)) %>%
  kbl(col.names = c("Model", "Distributional", "Hybrid"),
      format = "latex",
      booktabs = TRUE,
      caption = "AIC values for predicting relatedness judgments using Distributional Distance alone or both Distributional Distance and Sensorimotor Distance (AIC for Sensorimotor Distance was 2134.78.)",
      label = "aic") %>%
  kable_styling(latex_options = "hold_position")



best_layers %>%
  select(Model, aic_bert_vs_sm, aic_combined_vs_sm) %>%
  pivot_longer(cols = c(aic_bert_vs_sm, aic_combined_vs_sm),
               names_to = "type", values_to = "aic_diff") %>%
  mutate(type = ifelse(type == "aic_bert_vs_sm", "Distributional only", "Distributional + Sensorimotor"),
         type = factor(type, levels = c("Distributional only", "Distributional + Sensorimotor")),
         Model = fct_reorder(Model, aic_diff)) %>%
  ggplot(aes(x = aic_diff, y = Model, fill = type)) +
  geom_col(position = "dodge") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 4, linetype = "dotted", color = "gray50") +
  geom_vline(xintercept = -4, linetype = "dotted", color = "gray50") +
  labs(x = "ΔAIC vs. Sensorimotor baseline", 
       y = NULL,
       fill = NULL) +
  scale_fill_manual(values = c("Distributional only" = "gray60", 
                                "Distributional + Sensorimotor" = "steelblue")) +
  theme_minimal() +
  theme(text = element_text(size = 16),
        legend.position = "bottom")


# ============================================================
# 7. Summary
# ============================================================

best_layers %>%
  ungroup() %>%
  mutate(dist_better_than_sm = aic_bert_vs_sm < -4,
         hybrid_better_than_sm = aic_combined_vs_sm > 4,
         hybrid_better_than_dist = aic_combined_vs_dist > 4) %>%
  summarise(mean(dist_better_than_sm),
            mean(hybrid_better_than_sm),
            mean(hybrid_better_than_dist))



### Visualize raw AIC
aic_summary_full <- best_layers %>%
  select(Model, Layer, aic_sm, aic_bert, aic_combined) %>%
  pivot_longer(cols = c(aic_sm, aic_bert, aic_combined),
               names_to = "type", values_to = "aic") %>%
  mutate(type = case_when(
    type == "aic_sm" ~ "Sensorimotor",
    type == "aic_bert" ~ "Distributional",
    type == "aic_combined" ~ "Hybrid"
  ),
  type = factor(type, levels = c("Sensorimotor", "Distributional", "Hybrid")))

# Get sensorimotor baseline (should be the same for all models)
aic_sm_baseline <- aic_summary_full %>%
  filter(type == "Sensorimotor") %>%
  pull(aic) %>%
  unique()

# Filter to just distributional and hybrid
aic_summary_no_sm <- aic_summary_full %>%
  filter(type != "Sensorimotor")

aic_summary_se_no_sm <- aic_summary_no_sm %>%
  group_by(type) %>%
  summarize(
    mean_aic = mean(aic),
    se_aic = sd(aic) / sqrt(n())
  )

ggplot() +
  # Sensorimotor baseline as dashed line
  geom_hline(yintercept = aic_sm_baseline, 
             linetype = "dashed", color = "coral", linewidth = 1) +
  # Raw points for each model/layer
  geom_jitter(data = aic_summary_no_sm, 
              aes(x = type, y = aic, color = type),
              alpha = 0.3, width = 0.1, size = 2) +
  # Mean + SE
  geom_point(data = aic_summary_se_no_sm, 
             aes(x = type, y = mean_aic, color = type),
             size = 5) +
  geom_errorbar(data = aic_summary_se_no_sm, 
                aes(x = type, ymin = mean_aic - se_aic, 
                    ymax = mean_aic + se_aic, color = type), 
                width = 0.15, linewidth = 1.2) +
  annotate("text", x = 1.5, y = aic_sm_baseline + 30, 
           label = "Sensorimotor", color = "coral", size = 5)+
  labs(x = NULL, 
       y = "AIC (lower is better)",
       title = "Predicting Relatedness") +
  scale_color_manual(values = c("Distributional" = "gray60", 
                                "Hybrid" = "steelblue")) +
  theme_minimal() +
  theme(text = element_text(size = 16),
        legend.position = "none")

```




## Correlation between sensorimotor and distributional distance

```{r corr_dist_sm}
# Get best layer
df_by_layer = df_merged %>%
  group_by(Model, Multilingual, Layer, n_params) %>%
  summarise(r = cor(mean_relatedness, Distance, method = "pearson"),
            r2 = r ** 2,
            rho = cor(mean_relatedness, Distance, method = "spearman"),
            count = n())
df_best_layer <- df_by_layer %>%
  group_by(Model) %>%
  slice_max(r2, n = 1) %>%
  select(Model, Layer, r2)

# Filter for only best layers
df_best <- df_merged %>%
  semi_join(df_best_layer, by = c("Model", "Layer"))

# Pivot wider
df_wide <- df_best %>%
  mutate(`Sensorimotor Distance` = sensorimotor_distance) %>%
  select(word, sentence1, sentence2, `Sensorimotor Distance`, Model, Distance) %>%
  pivot_wider(names_from = Model, values_from = Distance)

# Compute correlation matrix
cor_matrix <- df_wide %>%
  select(`Sensorimotor Distance`, where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

# Plot the correlation matrix
ggcorrplot(cor_matrix, 
           hc.order = FALSE,
           method = "square") +
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10)
  )




```



# Baseline with LS Norms

```{r}
df_with_baseline = read_csv("../../data/processed/sentence_pairs_with_baseline.csv") %>%
  drop_na(sensorimotor_distance) %>%
  drop_na(baseline_distance)

nrow(df_with_baseline)
```


## Does sensorimotor distance predict relatedness above the baseline?

We also ask whether whether our measure of contextualized sensorimotor distance predicts relatedness above and beyond a baseline that simply considers the decontextualized Lancaster Sensorimotor Norms for the dismabiguating words in a sentence. (We find that it does.)

```{r}

model_bow_sm  = lmer(data = df_with_baseline,
                mean_relatedness ~ baseline_distance + sensorimotor_distance +
                  (1| word),
                REML = FALSE)

model_just_bow  = lmer(data = df_with_baseline,
                mean_relatedness ~ baseline_distance + 
                  (1| word),
                REML = FALSE)

model_just_sm  = lmer(data = df_with_baseline,
                mean_relatedness ~ sensorimotor_distance + 
                  (1| word),
                REML = FALSE)

summary(model_bow_sm)
anova(model_bow_sm, model_just_bow)
anova(model_bow_sm, model_just_sm)


```


# Predicting Trott & Bergen (2023)

Here, we also ask whether sensorimotor distance can account for variance in RT and accuracy on a primed sensibility judgment task.

```{r}
# ============================================================
# Load T&B 2023 data and merge with norms
# ============================================================

df_s1 <- read_csv("../../data/tb2023/polysemy_s1_final.csv")
df_s2 <- read_csv("../../data/tb2023/polysemy_s2_final.csv")
df_tb2023 <- bind_rows(df_s1, df_s2)
nrow(df_tb2023)

# Item-level aggregates
df_tb2023_agg_acc <- df_tb2023 %>%
  group_by(word, same, version) %>%
  summarise(accuracy = mean(correct_response), .groups = "drop")

df_tb2023_agg_rt <- df_tb2023 %>%
  filter(correct_response == TRUE) %>%
  group_by(word, same, version) %>%
  summarise(mean_rt = mean(rt), .groups = "drop") %>%
  mutate(mean_log_rt = log10(mean_rt))

# Merge with sensorimotor norms (for Part 1)
df_sm_acc <- df_rawc_with_norms %>%
  select(word, same, version, sensorimotor_distance,
         action_distance, perceptual_distance) %>%
  inner_join(df_tb2023_agg_acc, by = c("word", "same", "version"))

df_sm_rt <- df_rawc_with_norms %>%
  select(word, same, version, sensorimotor_distance,
         action_distance, perceptual_distance) %>%
  inner_join(df_tb2023_agg_rt, by = c("word", "same", "version"))

# Merge with full distributional sweep (for Part 2)
df_full_acc <- df_merged %>%
  inner_join(df_tb2023_agg_acc, by = c("word", "same", "version"))

df_full_rt <- df_merged %>%
  inner_join(df_tb2023_agg_rt, by = c("word", "same", "version"))

# Sanity check
nrow(df_sm_acc)
nrow(df_sm_rt)
nrow(df_full_acc)
nrow(df_full_rt)
```


## Part 1: SM on its own

```{r}
mod_sm_acc = lmer(data = df_sm_acc, 
                accuracy ~ sensorimotor_distance +
                  (1 | word))

summary(mod_sm_acc)


mod_sm_rt = lmer(data = df_sm_rt, 
                mean_log_rt ~ sensorimotor_distance +
                  (1 | word))

summary(mod_sm_rt)

```


## Part 2: Comopare to distributional distance

Here, we compare sensorimotor distance to the best-performing distributional distance measure from the RAW-C analysis.

```{r sm_dd_tb2023}

# --- Compute AIC for each model/layer: SM, Distributional, Hybrid ---

tb_acc_results <- df_full_acc %>%
  group_by(Model, Layer, n_params, Multilingual) %>%
  summarize(
    r2_sm       = summary(lm(accuracy ~ sensorimotor_distance))$r.squared,
    aic_sm      = AIC(lm(accuracy ~ sensorimotor_distance)),
    r2_bert     = summary(lm(accuracy ~ Distance))$r.squared,
    aic_bert    = AIC(lm(accuracy ~ Distance)),
    r2_combined = summary(lm(accuracy ~ Distance + sensorimotor_distance))$r.squared,
    aic_combined = AIC(lm(accuracy ~ Distance + sensorimotor_distance)),
    .groups = "drop"
  ) %>%
  mutate(
    aic_bert_vs_sm       = aic_bert - aic_sm,
    aic_combined_vs_sm   = aic_sm - aic_combined,
    aic_combined_vs_dist = aic_bert - aic_combined
  )

tb_rt_results <- df_full_rt %>%
  group_by(Model, Layer, n_params, Multilingual) %>%
  summarize(
    r2_sm       = summary(lm(mean_log_rt ~ sensorimotor_distance))$r.squared,
    aic_sm      = AIC(lm(mean_log_rt ~ sensorimotor_distance)),
    r2_bert     = summary(lm(mean_log_rt ~ Distance))$r.squared,
    aic_bert    = AIC(lm(mean_log_rt ~ Distance)),
    r2_combined = summary(lm(mean_log_rt ~ Distance + sensorimotor_distance))$r.squared,
    aic_combined = AIC(lm(mean_log_rt ~ Distance + sensorimotor_distance)),
    .groups = "drop"
  ) %>%
  mutate(
    aic_bert_vs_sm       = aic_bert - aic_sm,
    aic_combined_vs_sm   = aic_sm - aic_combined,
    aic_combined_vs_dist = aic_bert - aic_combined
  )

# --- Best layer per model (selected by best distributional R²) ---

best_layers_acc <- tb_acc_results %>%
  group_by(Model, n_params, Multilingual) %>%
  slice_max(r2_bert, n = 1) %>%
  ungroup()

best_layers_rt <- tb_rt_results %>%
  group_by(Model, n_params, Multilingual) %>%
  slice_max(r2_bert, n = 1) %>%
  ungroup()

print(best_layers_acc)
print(best_layers_rt)

# --- Summary counts ---

best_layers_acc %>%
  mutate(dist_better_than_sm     = aic_bert_vs_sm < -4,
         hybrid_better_than_sm   = aic_combined_vs_sm > 4,
         hybrid_better_than_dist = aic_combined_vs_dist > 4) %>%
  summarise(across(c(dist_better_than_sm,
                     hybrid_better_than_sm,
                     hybrid_better_than_dist), mean))

best_layers_rt %>%
  mutate(dist_better_than_sm     = aic_bert_vs_sm < -4,
         hybrid_better_than_sm   = aic_combined_vs_sm > 4,
         hybrid_better_than_dist = aic_combined_vs_dist > 4) %>%
  summarise(across(c(dist_better_than_sm,
                     hybrid_better_than_sm,
                     hybrid_better_than_dist), mean))

# --- LaTeX tables ---

best_layers_acc %>%
  select(Model, aic_bert, aic_combined) %>%
  mutate(across(starts_with("aic"), round)) %>%
  kbl(col.names = c("Model", "Distributional", "Hybrid"),
      format = "latex", booktabs = TRUE,
      caption = sprintf("AIC values for predicting accuracy (Trott \\& Bergen, 2023). AIC for Sensorimotor Distance alone was %.2f.",
                        unique(best_layers_acc$aic_sm)),
      label = "aic_tb_acc") %>%
  kable_styling(latex_options = "hold_position")

best_layers_rt %>%
  select(Model, aic_bert, aic_combined) %>%
  mutate(across(starts_with("aic"), round)) %>%
  kbl(col.names = c("Model", "Distributional", "Hybrid"),
      format = "latex", booktabs = TRUE,
      caption = sprintf("AIC values for predicting log RT (Trott \\& Bergen, 2023). AIC for Sensorimotor Distance alone was %.2f.",
                        unique(best_layers_rt$aic_sm)),
      label = "aic_tb_rt") %>%
  kable_styling(latex_options = "hold_position")

# ============================================================
# Figures: AIC dot plots (matching your existing style)
# ============================================================

# --- Accuracy ---

aic_long_acc <- best_layers_acc %>%
  select(Model, Layer, aic_sm, aic_bert, aic_combined) %>%
  pivot_longer(cols = c(aic_sm, aic_bert, aic_combined),
               names_to = "type", values_to = "aic") %>%
  mutate(type = case_when(
    type == "aic_sm"       ~ "Sensorimotor",
    type == "aic_bert"     ~ "Distributional",
    type == "aic_combined" ~ "Hybrid"
  ),
  type = factor(type, levels = c("Sensorimotor", "Distributional", "Hybrid")))

baseline_acc <- aic_long_acc %>% filter(type == "Sensorimotor") %>% pull(aic) %>% unique()

aic_long_acc_no_sm <- aic_long_acc %>% filter(type != "Sensorimotor")

aic_se_acc <- aic_long_acc_no_sm %>%
  group_by(type) %>%
  summarize(mean_aic = mean(aic),
            se_aic = sd(aic) / sqrt(n()),
            .groups = "drop")

# Dynamic offset for the label (5% of y-range above baseline)
offset_acc <- diff(range(aic_long_acc_no_sm$aic)) * 0.05

ggplot() +
  geom_hline(yintercept = baseline_acc,
             linetype = "dashed", color = "coral", linewidth = 1) +
  geom_jitter(data = aic_long_acc_no_sm,
              aes(x = type, y = aic, color = type),
              alpha = 0.3, width = 0.1, size = 2) +
  geom_point(data = aic_se_acc,
             aes(x = type, y = mean_aic, color = type),
             size = 5) +
  geom_errorbar(data = aic_se_acc,
                aes(x = type, ymin = mean_aic - se_aic,
                    ymax = mean_aic + se_aic, color = type),
                width = 0.15, linewidth = 1.2) +
  annotate("text", x = 1.5, y = baseline_acc + offset_acc,
           label = "Sensorimotor", color = "coral", size = 5) +
  labs(x = NULL, y = "AIC (lower is better)",
       title = "Predicting Accuracy") +
  scale_color_manual(values = c("Distributional" = "gray60",
                                "Hybrid" = "steelblue")) +
  theme_minimal() +
  theme(text = element_text(size = 16), legend.position = "none")

# --- RT ---

aic_long_rt <- best_layers_rt %>%
  select(Model, Layer, aic_sm, aic_bert, aic_combined) %>%
  pivot_longer(cols = c(aic_sm, aic_bert, aic_combined),
               names_to = "type", values_to = "aic") %>%
  mutate(type = case_when(
    type == "aic_sm"       ~ "Sensorimotor",
    type == "aic_bert"     ~ "Distributional",
    type == "aic_combined" ~ "Hybrid"
  ),
  type = factor(type, levels = c("Sensorimotor", "Distributional", "Hybrid")))

baseline_rt <- aic_long_rt %>% filter(type == "Sensorimotor") %>% pull(aic) %>% unique()

aic_long_rt_no_sm <- aic_long_rt %>% filter(type != "Sensorimotor")

aic_se_rt <- aic_long_rt_no_sm %>%
  group_by(type) %>%
  summarize(mean_aic = mean(aic),
            se_aic = sd(aic) / sqrt(n()),
            .groups = "drop")

offset_rt <- diff(range(aic_long_rt_no_sm$aic)) * 0.05

ggplot() +
  geom_hline(yintercept = baseline_rt,
             linetype = "dashed", color = "coral", linewidth = 1) +
  geom_jitter(data = aic_long_rt_no_sm,
              aes(x = type, y = aic, color = type),
              alpha = 0.3, width = 0.1, size = 2) +
  geom_point(data = aic_se_rt,
             aes(x = type, y = mean_aic, color = type),
             size = 5) +
  geom_errorbar(data = aic_se_rt,
                aes(x = type, ymin = mean_aic - se_aic,
                    ymax = mean_aic + se_aic, color = type),
                width = 0.15, linewidth = 1.2) +
  annotate("text", x = 1.5, y = baseline_rt + offset_rt,
           label = "Sensorimotor", color = "coral", size = 5) +
  labs(x = NULL, y = "AIC (lower is better)",
       title = "Predicting RT") +
  scale_color_manual(values = c("Distributional" = "gray60",
                                "Hybrid" = "steelblue")) +
  theme_minimal() +
  theme(text = element_text(size = 16), legend.position = "none")
```

## Part 3: Does SM account for sense boundaries?

```{r}

# --- Compute AIC for each model/layer: same alone vs. same + continuous ---

tb_acc_q3 <- df_full_acc %>%
  group_by(Model, Layer, n_params, Multilingual) %>%
  summarize(
    aic_same      = AIC(lm(accuracy ~ same)),
    aic_same_sm   = AIC(lm(accuracy ~ same + sensorimotor_distance)),
    aic_same_dist = AIC(lm(accuracy ~ same + Distance)),
    aic_same_both = AIC(lm(accuracy ~ same + Distance + sensorimotor_distance)),
    r2_same_dist  = summary(lm(accuracy ~ same + Distance))$r.squared,
    .groups = "drop"
  ) %>%
  mutate(
    delta_aic_sm   = aic_same - aic_same_sm,    # positive = SM adds over same
    delta_aic_dist = aic_same - aic_same_dist,  # positive = Distributional adds over same
    delta_aic_both = aic_same - aic_same_both   # positive = both add over same
  )

tb_rt_q3 <- df_full_rt %>%
  group_by(Model, Layer, n_params, Multilingual) %>%
  summarize(
    aic_same      = AIC(lm(mean_log_rt ~ same)),
    aic_same_sm   = AIC(lm(mean_log_rt ~ same + sensorimotor_distance)),
    aic_same_dist = AIC(lm(mean_log_rt ~ same + Distance)),
    aic_same_both = AIC(lm(mean_log_rt ~ same + Distance + sensorimotor_distance)),
    r2_same_dist  = summary(lm(mean_log_rt ~ same + Distance))$r.squared,
    .groups = "drop"
  ) %>%
  mutate(
    delta_aic_sm   = aic_same - aic_same_sm,
    delta_aic_dist = aic_same - aic_same_dist,
    delta_aic_both = aic_same - aic_same_both
  )

# --- Best layer per model (selected by best same + Distributional R²) ---

best_layers_acc_q3 <- tb_acc_q3 %>%
  group_by(Model, n_params, Multilingual) %>%
  slice_max(r2_same_dist, n = 1) %>%
  ungroup()

best_layers_rt_q3 <- tb_rt_q3 %>%
  group_by(Model, n_params, Multilingual) %>%
  slice_max(r2_same_dist, n = 1) %>%
  ungroup()

print(best_layers_acc_q3)
print(best_layers_rt_q3)

# --- Summary counts: how often does each augmentation add over same alone ---

best_layers_acc_q3 %>%
  mutate(sm_adds   = delta_aic_sm > 4,
         dist_adds = delta_aic_dist > 4,
         both_add  = delta_aic_both > 4) %>%
  summarise(across(c(sm_adds, dist_adds, both_add), mean))

best_layers_rt_q3 %>%
  mutate(sm_adds   = delta_aic_sm > 4,
         dist_adds = delta_aic_dist > 4,
         both_add  = delta_aic_both > 4) %>%
  summarise(across(c(sm_adds, dist_adds, both_add), mean))


make_aic_plot_q3 <- function(best_layers_df, title) {
  aic_long <- best_layers_df %>%
    select(Model, Layer, aic_same, aic_same_sm, aic_same_dist, aic_same_both) %>%
    pivot_longer(cols = c(aic_same, aic_same_sm, aic_same_dist, aic_same_both),
                 names_to = "type", values_to = "aic") %>%
    mutate(type = case_when(
      type == "aic_same"      ~ "Same only",
      type == "aic_same_sm"   ~ "Same + SM",
      type == "aic_same_dist" ~ "Same + Distributional",
      type == "aic_same_both" ~ "Same + Both"
    ),
    type = factor(type, levels = c("Same only", "Same + SM",
                                    "Same + Distributional", "Same + Both")))

  baseline <- aic_long %>% filter(type == "Same only") %>% pull(aic) %>% unique()

  aic_long_no_base <- aic_long %>% filter(type != "Same only")

  aic_se <- aic_long_no_base %>%
    group_by(type) %>%
    summarize(mean_aic = mean(aic),
              se_aic = sd(aic) / sqrt(n()),
              .groups = "drop")

  offset <- diff(range(aic_long_no_base$aic)) * 0.05

  ggplot() +
    geom_hline(yintercept = baseline,
               linetype = "dashed", color = "coral", linewidth = 1) +
    geom_jitter(data = aic_long_no_base,
                aes(x = type, y = aic, color = type),
                alpha = 0.3, width = 0.1, size = 2) +
    geom_point(data = aic_se,
               aes(x = type, y = mean_aic, color = type),
               size = 5) +
    geom_errorbar(data = aic_se,
                  aes(x = type, ymin = mean_aic - se_aic,
                      ymax = mean_aic + se_aic, color = type),
                  width = 0.15, linewidth = 1.2) +
    annotate("text", x = 2, y = baseline + offset,
             label = "Same only", color = "coral", size = 5) +
    labs(x = NULL, y = "AIC (lower is better)", title = title) +
    scale_color_manual(values = c("Same + SM" = "coral",
                                  "Same + Distributional" = "gray60",
                                  "Same + Both" = "steelblue")) +
    theme_minimal() +
    theme(text = element_text(size = 16), legend.position = "none",
          axis.text.x = element_text(angle = 20, hjust = 1))
}

make_aic_plot_q3(best_layers_acc_q3, "Predicting Accuracy")
make_aic_plot_q3(best_layers_rt_q3, "Predicting RT")
```


# Comparison to Trott (2024) LLM Norms

Here, we conduct a more in-depth comparison to the GPT-4-generated norms from Trott (2024).

## Load data

```{r}
df_gpt_perception = read_csv("../../data/trott2024/cs_norms_perception_gpt-4.csv")
df_gpt_action = read_csv("../../data/trott2024/cs_norms_action_gpt-4.csv")

df_both_gpt = df_gpt_perception %>%
  inner_join(df_gpt_action)

df_gpt_with_human = df_both_gpt %>%
  inner_join(df_contextualized_meanings)



### RAW-C With GPT and human SM distance
df_rawc_with_sm = read_csv("../../data/trott2024/rawc_with_sm.csv")

```

## Dimension-level correlation

```{r gpt_corrs}
df_summ = df_gpt_with_human %>%
  summarise(Vision = cor(Vision.M, Vision, method = "spearman"),
            Hearing = cor(Hearing.M, Hearing, method = "spearman"),
            Touch = cor(Touch.M, Touch, method = "spearman"),
            Olfaction = cor(Olfaction.M, Olfaction, method = "spearman"),
            Taste = cor(Taste.M, Taste, method = "spearman"),
            Interoception = cor(Interoception.M, Interoception, method = "spearman"),
            Hand_arm = cor(Hand_arm.M, Hand_arm, method = "spearman"),
            Foot_leg = cor(Foot_leg.M, Foot_leg, method = "spearman"),
            Head = cor(Head.M, Head, method = "spearman"),
            Torso = cor(Torso.M, Torso, method = "spearman"),
            Mouth_throat = cor(Mouth_throat.M, Mouth_throat, method = "spearman"))

df_summ

df_long = df_summ %>%
  pivot_longer(everything(), names_to = "Factor", values_to = "Correlation")

df_long %>%
  ggplot(aes(x = reorder(Factor, Correlation), y = Correlation)) +
  geom_bar(stat = "identity") +
  labs(x = "Factor", y = "Correlation") +
  theme_minimal()
```



## Compare correlation matrices

```{r gpt_corrs_comparison}

library(tidyverse)
library(ggcorrplot)
library(patchwork)

dim_order <- c("Vision", "Hearing", "Olfaction", "Taste", "Interoception", 
               "Touch", "Mouth_throat", "Head", "Torso", "Hand_arm", "Foot_leg")

# Human matrix: .M columns
human_cols <- df_gpt_with_human %>%
  select(all_of(paste0(dim_order, ".M"))) %>%
  rename_with(~ str_remove(., "\\.M"))
cors_human <- cor(human_cols, use = "pairwise.complete.obs")

# GPT matrix: bare-named columns
gpt_cols <- df_gpt_with_human %>%
  select(all_of(dim_order))
cors_gpt <- cor(gpt_cols, use = "pairwise.complete.obs")

# Difference (GPT - Human)
cors_diff <- cors_gpt - cors_human

ggcorrplot(cors_human, hc.order = FALSE, type = "upper",
                      lab = TRUE, lab_size = 2.5) +
  ggtitle("Human") +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))

ggcorrplot(cors_gpt, hc.order = FALSE, type = "upper",
                    lab = TRUE, lab_size = 2.5) +
  ggtitle("GPT-4") +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))

ggcorrplot(cors_diff, hc.order = FALSE, type = "upper",
                     lab = TRUE, lab_size = 2.5,
                     colors = c("#2166AC", "white", "#B2182B")) +
  ggtitle("GPT − Human") +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))



### Statistical comparison
library(vegan)


# Mantel test
mantel_result <- mantel(cors_human, cors_gpt, 
                        method = "pearson",   # or "spearman"
                        permutations = 9999)

mantel_result
mantel_result <- mantel(cors_human, cors_gpt, 
                        method = "spearman",   # or "spearman"
                        permutations = 9999)

mantel_result


## Or just straight-up correlation
upper_human <- cors_human[upper.tri(cors_human)]
upper_gpt   <- cors_gpt[upper.tri(cors_gpt)]

# Correlation between the two
cor(upper_human, upper_gpt, method = "pearson")
cor(upper_human, upper_gpt, method = "spearman")

```

## Substitution analyses


### Homonymy vs. Polysemy

Here, we ask whether homonyms show bigger sensorimotor distance for GPT-generated norms as well.

```{r}

model_full = lmer(data = df_rawc_with_sm,
                  sm_gpt ~ same +
                    (1 + same | word),
                  control=lmerControl(optimizer="bobyqa"),
                  REML = FALSE)

model_reduced = lmer(data = df_rawc_with_sm,
                  sm_gpt ~ # same +
                    (1 + same | word),
                  control=lmerControl(optimizer="bobyqa"),
                  REML = FALSE)

summary(model_full)
anova(model_full, model_reduced)


model_full_at = lmer(data = df_rawc_with_sm,
                  sm_gpt ~ same * ambiguity_type +
                    (1 + same | word),
                  control=lmerControl(optimizer="bobyqa"),
                  REML = FALSE)

model_reduced_at = lmer(data = df_rawc_with_sm,
                  sm_gpt ~ same + ambiguity_type +
                    (1 + same | word),
                  control=lmerControl(optimizer="bobyqa"),
                  REML = FALSE)

summary(model_full_at)
anova(model_full_at, model_reduced_at)


df_rawc_with_sm %>%
  group_by(same, ambiguity_type) %>%
  summarise(mean_gpt = mean(sm_gpt),
            sd_gpt = sd(sm_gpt),
            mean_human = mean(sm_human),
            sd_human = sd(sm_human))



df_rawc_with_sm = df_rawc_with_sm %>%
  mutate(Same = case_when(
    same == TRUE ~ "Same Sense",
    same == FALSE ~ "Different Sense"
  ))




df_rawc_with_sm %>%
  ggplot(aes(x = sm_gpt,
             y = ambiguity_type,
             fill = Same)) +
  geom_density_ridges2(aes(height = ..density..), 
                       color = NA,
                       alpha = 0.5, 
                       scale=0.85, 
                       stat="density") +
  labs(x = "Sensorimotor Distance (GPT-4)",
       y = NULL,
       fill = NULL) +
  scale_fill_manual(
    values = c("Same Sense" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[1],
               "Different Sense" = viridisLite::viridis(2, option = "mako", begin = 0.8, end = 0.15)[2])
  ) + 
  theme_minimal() +
  theme(text = element_text(size = 20))



```


### Dominance

We also replicate the dominance analysis.

```{r gpt_sm_strength}

df_dom_gpt = df_gpt_with_human %>%
  inner_join(df_dominance_individual)


df_dom_gpt <- df_dom_gpt %>%
  inner_join(df_lancaster)

nrow(df_dom_gpt)


df_dom_gpt = df_dom_gpt %>%
  rowwise() %>%
  mutate(max_strength_gpt = max(
    c(
      ## Modalities
      Vision,
      Hearing,
      Olfaction,
      Touch,
      Taste, 
      Interoception,
      ## Effectors
      Head,
      Mouth_throat,
      Torso,
      Hand_arm,
      Foot_leg
    )
  ),
  minkowski3_strength_gpt = sum(c(Vision,
      Hearing,
      Olfaction,
      Touch,
      Taste, 
      Interoception,
      ## Effectors
      Head,
      Mouth_throat,
      Torso,
      Hand_arm,
      Foot_leg)^3)^(1/3),
max_strength = max(
    c(
      ## Modalities
      Vision.M,
      Hearing.M,
      Olfaction.M,
      Touch.M,
      Taste.M, 
      Interoception.M,
      ## Effectors
      Head.M,
      Mouth_throat.M,
      Torso.M,
      Hand_arm.M,
      Foot_leg.M
    )
  ),
  minkowski3_strength = sum(c(Vision.M, Hearing.M, Olfaction.M, Touch.M, Taste.M,
                               Interoception.M, Head.M, Mouth_throat.M, Torso.M,
                               Hand_arm.M, Foot_leg.M)^3)^(1/3)
  ) %>%
  ungroup()



cor.test(df_dom_gpt$max_strength_gpt, df_dom_gpt$max_strength)
cor.test(df_dom_gpt$minkowski3_strength_gpt, df_dom_gpt$minkowski3_strength)

df_dom_gpt %>%
  ggplot(aes(x = minkowski3_strength_gpt,
             y = minkowski3_strength)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm") +
  labs(y = "Minkowski-3 Strength (GPT-4)",
       x = "Minkowski-3 Strength (Human)") +
  theme_minimal() +
  theme(text = element_text(size = 16), legend.position = "none")


```


### Does GPT sensorimotor strength predict dominance?


```{r gpt_dominance}

  
df_dom_gpt %>%
  ggplot(aes(x = minkowski3_strength,
             y = dominance)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "lm") +
  labs(x = "Minkowski-3 Sensorimotor Strength (GPT-4)",
       y = "Dominance") +
  theme_minimal() +
  theme(text = element_text(size=16))

model_full = lmer(data = df_dom_gpt,
                dominance ~ 
                  max_strength_gpt +
                  Max_strength.sensorimotor + Minkowski3.sensorimotor +
                  (1 | word),
                REML = FALSE)

model_reduced = lmer(data = df_dom_gpt,
                dominance ~ 
                  # max_strength_gpt + 
                  Max_strength.sensorimotor + Minkowski3.sensorimotor +
                  (1 | word),
                REML = FALSE)
summary(model_full)
anova(model_full, model_reduced)




model_full = lmer(data = df_dom_gpt,
                dominance ~ 
                  minkowski3_strength_gpt +
                  Max_strength.sensorimotor + Minkowski3.sensorimotor +
                  (1 | word),
                REML = FALSE)

model_reduced = lmer(data = df_dom_gpt,
                dominance ~ 
                  # minkowski3_strength_gpt + 
                  Max_strength.sensorimotor + Minkowski3.sensorimotor +
                  (1 | word),
                REML = FALSE)
summary(model_full)
anova(model_full, model_reduced)

```

